Doublet splitting of fusion alpha-particle driven ion cyclotron emission

### 2022-07-23 manuscript.tex
further improvements based off referee comments

The effectiveness of analytical theory to reproduce the observed cyclotron
harmonic peaks in the fluctuations of the magnetic field was first explained
somewhat by 1D PIC simulations. In these simulations the onset of nonlinearity
appeared to imprint the amplitudes of each of the harmonics into the signal as
they are at the end of the linear phase; each mode grows independently until 
growth is terminated for all modes when the fastest mode enters its nonlinear
phase meaning that both the amount of free energy available for instability has
decreased and that the unstable energy ion distribution function has been
perturbed into a more stable state. Analysis shows that only higher cyclotron
harmonics are destabilised\cite{McClements2015} to the MCI, however experimental
observations clearly indicate power at theoretically stable lower
harmonics\cite{Cottrell1988}. In a further step towards accurate recreation of
experimental observations, nonlinear phases of hybrid\cite{Carbajal2014} and
fully kinetic\cite{Cook2013,Chapman2017} PIC simulations went on to show that
wave-wave coupling was responsible for raising the amplitude of lower cyclotron
harmonics. Most recently, hybrid 2D simulations have begun to shed light on the
2D3V dynamics of ICE\cite{Carbajal2021}.

@@

Although curvature and grad-B drift terms would be present in fast
alpha-particle populations on JET, it is worth noting that they are both of
order $\rho^*$ and have been neglected in these calculations where $u_\bot$ is
always large.

@@

@@

Doublet splitting can be seen in Fig. \ref{fig:2D_imag_physicalpitch}, which is the
counterpart figure to Fig. \ref{fig:2D_F12_physicalpitch} as Fig.
\ref{fig:2D_imag_zeropitch} is to Fig. \ref{fig:2D_F12_zeropitch},
at e.g. the $12^{th}$ harmonic  near $(k_\parallel \simeq 0\; \Omega_{i}/V_A,
k_\bot \simeq 13\; \Omega_{i}/V_A)$, where $V_A$ is the Alfv{\'e}n speed.
Other possible mechanisms for doublet splitting could be grad-B or curvature
drifts\cite{Cottrell1993,Fulop1998} or potentially any mechanism that is capable
of modifying the wave-particle resonance condition.
Qualitatively, scattered points in Fig.
\ref{fig:2D_F12_physicalpitch}(a) resembles the experimental observations given
by the black trace in that the
signal below the $7^{th}$ cyclotron harmonics is lower and flatter with well
defined peaks and the signal beyond the $7^{th}$ harmonic is a step-change
larger and the harmonics are less easily identifiable. These features cannot be
be recreated with a 1D calculation for these distributions functions. Note that
Fig. \ref{fig:2D_imag_physicalpitch} shows these two different regimes more
clearly. The lower frequency behaviour, where $k_\perp < 7 \Omega_i/V_A$, is
characterised by narrow regions of instability near Doppler shifted cyclotron
harmonics that are stable when $k_\parallel=0$ and unstable at larger
negative parallel wavenumbers. The higher frequency behaviour in the region
$\omega\geq 7\Omega_i$ is dominated by broader regions of instability in
frequency and wavenumber near to Doppler shifted cyclotron harmonics, which have
a complex structure in wavenumber space. Modes numbering 1, 2, 6, 7, 8, 12, and
13 are stable when propagating exactly perpendicular whilst obliquely
propagating modes remain unstable. The magnitude of the Doppler shift of
the growing modes can reach $\approx \Omega_i$ for large propagation angles at 
lower frequencies and also for
oblique angles at higher frequencies; see, for example, the mode
associated with the $15^{th}$ cyclotron harmonic near $k_\parallel \simeq
-\frac{3}{4} \Omega_i/V_A, k_\perp \simeq 14 \Omega_i/V_A$, i.e. a propagation
angle of $93^\circ$. Furthermore, the breadth of the harmonics can be
attributed to the contributions from multiple propagation angles even though the
lower harmonics are largely stable for forward travelling and perpendicular
propagation angles.

\begin{figure}[ht!]
    % \centering
    \raggedleft
\includegraphics[scale=0.9]{fig3.pdf}
    \caption{(colour online) The imaginary component of the frequencies for the
    case with parallel drift $u_{\parallel} = -0.646 u_0 \simeq -1.1 V_A$.
    Panel (a) plots growth rate against frequency where shading indicates
    parallel wavenumber in units of $\Omega_i/V_A$ and the black trace shows
    the re-scaled ICE signal from JET Pulse \#21648 digitised from Fig. 2 of
    Cottrell et al 1993 (with permission from IOP Publishing), (b) plots the solutions
    in the $(\omega,k_\parallel)$ plane where shading indicates growth rate. Note
    that the Doppler shift associated with the non-zero parallel drift speed
    broadens the unstable peaks in frequency. The $10^{th}$ and $11^{th}$ cyclotron
    harmonics are split into doublets, i.e. they have two peaks.}
    \label{fig:2D_F12_physicalpitch}
\end{figure}

@@

>>>>

The effectiveness of analytical theory to reproduce the observed cyclotron
harmonic peaks in the fluctuations of the magnetic field was first explained
somewhat by 1D PIC simulations. In these simulations the onset of nonlinearity
appeared to imprint the amplitudes of each of the harmonics into the signal as
they are at the end of the linear phase; each mode grows independently until 
growth is terminated for all modes when the fastest mode enters its nonlinear
phase meaning that both the amount of free energy available for instability has
decreased and that the unstable energy ion distribution function has been
perturbed into a more stable state. Analysis shows that only higher cyclotron
harmonics are destabilised\cite{McClements2015} to the MCI under these
conditions, however experimental
observations clearly indicate power at theoretically stable lower
harmonics\cite{Cottrell1988}. In a further step towards accurate recreation of
experimental observations, nonlinear phases of hybrid\cite{Carbajal2014} and
fully kinetic\cite{Cook2013,Chapman2017} PIC simulations went on to show that
wave-wave coupling was responsible for raising the amplitude of lower cyclotron
harmonics. Most recently, hybrid 2D simulations have begun to shed light on the
2D3V dynamics of ICE\cite{Carbajal2021}.

Finer detailed structure has been obvserved in ICE spectra, such as doublet
splitting of individual harmonics on JET\cite{Cottrell1993} and
DIII-D\cite{Thome2018} and even
multiplet splitting on TUMAN-3M\cite{Askinazi2018} on EAST\cite{Lui2020}.
This paper highlights one underlying kinetic mechinism that produces
doublet splitting due to Doppler shifted resonances. In general,
splitting could be produced by grad-B or curvature
drifts\cite{Cottrell1993,Fulop1998} or potentially any mechanism that is capable
of adding a Doppler shift to the wave-particle resonance condition.

@@

@@

Although curvature and grad-B drift terms would be present in fast
alpha-particle populations on JET, it is worth noting that they are both of
order $\rho^*$ and have been neglected in these calculations where $u_\bot \gg
\rho^* u_0$ is always the case.

@@

Doublet splitting can be seen in Fig. \ref{fig:2D_imag_physicalpitch}, which is the
counterpart figure to Fig. \ref{fig:2D_F12_physicalpitch} as Fig.
\ref{fig:2D_imag_zeropitch} is to Fig. \ref{fig:2D_F12_zeropitch},
at e.g. the $12^{th}$ harmonic  near $(k_\parallel \simeq 0\; \Omega_{i}/V_A,
k_\bot \simeq 13\; \Omega_{i}/V_A)$, where $V_A$ is the Alfv{\'e}n speed.
Qualitatively, scattered points in Fig.
\ref{fig:2D_F12_physicalpitch}(a) resembles the experimental observations given
by the black trace in that the
signal below the $7^{th}$ cyclotron harmonics is lower and flatter with well
defined peaks and the signal beyond the $7^{th}$ harmonic is a step-change
larger and the harmonics are less easily identifiable. These features cannot be
be recreated with a 1D calculation for these distributions functions. Note that
Fig. \ref{fig:2D_imag_physicalpitch} shows these two different regimes more
clearly. The lower frequency behaviour, where $k_\perp < 7 \Omega_i/V_A$, is
characterised by narrow regions of instability near Doppler shifted cyclotron
harmonics that are stable when $k_\parallel=0$ and unstable at larger
negative parallel wavenumbers. The higher frequency behaviour in the region
$\omega\geq 7\Omega_i$ is dominated by broader regions of instability in
frequency and wavenumber near to Doppler shifted cyclotron harmonics, which have
a complex structure in wavenumber space. Modes numbering 1, 2, 6, 7, 8, 12, and
13 are stable when propagating exactly perpendicular whilst obliquely
propagating modes remain unstable. The magnitude of the Doppler shift of
the growing modes can reach $\approx \Omega_i$ for large propagation angles at 
lower frequencies and also for
oblique angles at higher frequencies; see, for example, the mode
associated with the $15^{th}$ cyclotron harmonic near $k_\parallel \simeq
-\frac{3}{4} \Omega_i/V_A, k_\perp \simeq 14 \Omega_i/V_A$, i.e. a propagation
angle of $93^\circ$. Furthermore, the breadth of the harmonics can be
attributed to the contributions from multiple propagation angles even though the
lower harmonics are largely stable for forward travelling and perpendicular
propagation angles.

\begin{figure}[ht!]
    % \centering
    \raggedleft
\includegraphics[scale=0.9]{fig3.pdf}
    \caption{(colour online) The imaginary component of the frequencies for the
    case with parallel drift $u_{\parallel} = -0.646 u_0 \simeq -1.1 V_A$.
    Panel (a) plots growth rate against frequency where shading indicates
    parallel wavenumber in units of $\Omega_i/V_A$ and the black trace shows
    the re-scaled ICE signal from JET Pulse \#26148 digitised from Fig. 2 of
    Cottrell et al 1993 (with permission from IOP Publishing), (b) plots the solutions
    in the $(\omega,k_\parallel)$ plane where shading indicates growth rate. Note
    that the Doppler shift associated with the non-zero parallel drift speed
    broadens the unstable peaks in frequency. The $10^{th}$ and $11^{th}$ cyclotron
    harmonics are split into doublets, i.e. they have two peaks.}
    \label{fig:2D_F12_physicalpitch}
\end{figure}

@@

\bibitem{Thome2018}
Thome K E \textit{et al} 2018 \textit{Rev. Sci. Instrum.} \textbf{89} 10I102

\bibitem{Askinazi2018}
Askinazi L G  \textit{et al} 2018 \textit{Nucl. Fusion} \textbf{58} 082003
